---
title: "Lecture 10: Point Level Models - Model Fitting, cont.."
output:
  revealjs::revealjs_presentation:
    theme: night
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(ggplot2)
library(dplyr)
library(mnormt)
library(gstat)
library(sp)
```

# Class Intro

## Intro Questions 
- Discuss how the spatial range could be extended to $\mathcal{R}^2$ rather than $\mathcal{R}^1$ in the Gaussian Process setting. What changes in this situation?
- For Today:
    - More Model Fitting


## Universal Kriging
- When covariate information is available for inclusion in the analysis, this is often referred to as *universal kriging*
- Now we have $$\boldsymbol{Y} = X \boldsymbol{\beta} + \boldsymbol{\epsilon}, \; \; \text{ where } \boldsymbol{\epsilon} \sim N(\boldsymbol{0}, \Sigma)$$
- The conditional distributions are very similar to what we have derived above, watch for HW question.

- In each case, kriging or universal kriging, it is still necessary to estimate the following parameters: $\sigma^2,$ $\tau^2$, $\phi$, and $\mu$ or $\beta$.
- This can be done with least-squares methods or in a Bayesian framework.


# Parameter Estimation

## About those other parameters

We still need

- to choose an appropriate covariance function (or semivariogram)
- and estimate parameters in that function


## Functions in R for 2D kriging
- The `krige` function in `gstat` contains a function for kriging ; however, again this requires a known variogram.

```{r, eval = F, echo = T }
data(meuse)
coordinates(meuse) = ~x +y
data(meuse.grid)
gridded(meuse.grid) = ~x + y
m <- vgm(.59, "Sph", 874, .04)
# ordinary kriging:
x <- krige(log(zinc)~1, meuse, meuse.grid, model = m)
spplot(x["var1.pred"], main = "ordinary kriging predictions")
```

## Functions in R for 2D kriging
- The `krige` function in `gstat` contains a function for kriging ; however, again this requires a known variogram.

```{r, eval = T, message = F}
data(meuse)
coordinates(meuse) = ~x +y
data(meuse.grid)
gridded(meuse.grid) = ~x + y
m <- vgm(.59, "Sph", 874, .04)
# ordinary kriging:
x <- krige(log(zinc)~1, meuse, meuse.grid, model = m)
spplot(x["var1.pred"], main = "ordinary kriging predictions")
```

## 